\mpx{}: Mixed Precision Training for JAX

### 2025-10-26 docs/paper/main.tex
updated readme and examples to reflect that many operations do not need to be forced to full precision as they already do this internally
added a small playground script for fp8 computations

\documentclass[copyright, logo]{dsme}

@@

For casting entire functions, \mpx{} provides the transformation \texttt{cast\_function(func, dtype, return\_dtype=None)}. 
This transformation returns a new function that first applies the specified input data type to all inputs (via \texttt{cast\_tree(tree, dtype)}), then calls the original function, and finally casts the outputs to the specified (optional) return data type. 
Furthermore, \mpx{} includes \texttt{force\_full\_precision(func, return\_dtype)}, which ensures computations are carried out in full precision. 
This feature is particularly important for operations that are prone to overflow in \texttt{float16}, such as \texttt{sum}, \texttt{mean}, or \texttt{softmax}.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \includegraphics[width=0.99\textwidth]{assets/EQXOverview.pdf}
        \caption{Full Precision Training via Equinox}
        \vspace{0.2cm}
    \end{subfigure}\hspace{0.1cm}
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \includegraphics[width=0.99\textwidth]{assets/MPXOverview.pdf}
        \caption{Half Precision Training via Equinox and \mpx{}}
    \end{subfigure}\hspace{0.1cm}
    \caption{Transformation of training pipelines via MPX. \capt{An existing training pipeline (Top figure) can be transformed, by applying \texttt{@mpx.filter\_grad} to the forward pass. It automatically casts all inputs to half precision and calculates the gradient via dynamic loss scaling and \texttt{eqx.filter\_grad}.}}
    \label{fig:overview}
\end{figure}

@@

\subsection{Model Implementation}

\begin{example}[t]
\begin{pbox}[]{}
\begin{minted}[linenos, breaklines, mathescape, fontsize=\footnotesize, xleftmargin=2em]{python}
class MultiHeadAttentionBlock(eqx.Module):
    dense_qs: eqx.nn.Linear
    dense_ks: eqx.nn.Linear
    dense_vs: eqx.nn.Linear
    dense_o: eqx.nn.Linear
    num_heads: int
    layer_norm: eqx.nn.LayerNorm

    def __init__(self, feature_dim, num_heads, key):
        self.num_heads = num_heads
        key, subkey = jax.random.split(key)
        self.dense_qs = eqx.nn.Linear(
            feature_dim, feature_dim, key=subkey)
        # same for dense_ks, dense_vs, dense_o

        self.layer_norm = eqx.nn.LayerNorm(feature_dim)

    def attention(q, k, v):
        attention_scores = q @ k.T / jnp.sqrt(q.shape[-1])
        attention_scores = mpx.force_full_precision(
            jax.nn.softmax, attention_scores.dtype)(attention_scores, axis=-1)
        return attention_scores @ v

    def __call__(self, inputs):
        inputs_after_layernorm = jax.vmap(mpx.force_full_precision(
            self.layer_norm, inputs.dtype))(inputs)
        qs = jax.vmap(self.dense_qs)(inputs_after_layernorm)
        qs = es.jax_einshape("n(hf)->hnf", qs, h=self.num_heads)
        # same for ks and vs...

        outputs = jax.vmap(self.attention, in_axes=(0, 0, 0))(qs, ks, vs)
        outputs = es.jax_einshape("hnf->n(hf)", outputs)
        outputs = jax.vmap(self.dense_o)(outputs)
        outputs += inputs

        return outputs
\end{minted}
\end{pbox}
\caption{Implementation of multi-headed self-attention via MPX. \capt{For the largest part, the model can be implemented in common toolboxes (here, Equinox~\citep{kidger2021equinox}). Only critical function that should be implemented in full precision, like softmax or layernorm (due to mean and std calculations), have to be transformed via \mpx{}.}}
\label{ex:model}
\end{example}

For the largest part, the implementation of the model must not be changed.
As \mpx{} works with arbitrary PyTrees, every Toolbox that defines their model/parameters as PyTrees, like Flax~\citep{flax2020github} or Equinox~\citep{kidger2021equinox} can be used in conjunction with \mpx{} (Example~\ref{ex:model}). 
Only parts of the model that contain critical operations with a high risk of overflows, like sum or mean, must be transformed to full precision via \texttt{mpx.force\_full\_precision}.

\subsection{Model Updates}
\label{sec:eqxvsmpxmodelupdate}
Model updates via \mpx{} closely resemble those performed with Equinox (Example~\ref{ex:update}). 
Instead of calling \texttt{eqx.filter\_grad}, \texttt{mpx.filter\_grad} is used. Similarly, \texttt{mpx.optimizer\_update} replaces invoking the Optax optimizer followed by a model update.
The remainder of the training step—such as loading batches, sharding the model and batch—remains unchanged. 
All these steps can be compiled using tools like \texttt{eqx.filter\_jit}.

>>>>

\documentclass[logo]{dsme}
% \documentclass[twocolumn]{article}

@@

For casting entire functions, \mpx{} provides the transformation \texttt{cast\_function(func, dtype, return\_dtype=None)}. 
This transformation returns a new function that first applies the specified input data type to all inputs (via \texttt{cast\_tree(tree, dtype)}), then calls the original function, and finally casts the outputs to the specified (optional) return data type. 
Furthermore, \mpx{} includes \texttt{force\_full\_precision(func, return\_dtype)}, which ensures computations are carried out in full precision. 
This feature is particularly important for operations that are prone to overflow in \texttt{float16}.

@@

% \begin{figure}[t]
%     \centering
%     \begin{subfigure}[t]{0.99\textwidth}
%         \centering
%         \includegraphics[width=0.99\textwidth]{assets/EQXOverview.pdf}
%         \caption{Full Precision Training via Equinox}
%         \vspace{0.2cm}
%     \end{subfigure}\hspace{0.1cm}
%     \begin{subfigure}[t]{0.99\textwidth}
%         \centering
%         \includegraphics[width=0.99\textwidth]{assets/MPXOverview.pdf}
%         \caption{Half Precision Training via Equinox and \mpx{}}
%     \end{subfigure}\hspace{0.1cm}
%     \caption{Transformation of training pipelines via MPX. \capt{An existing training pipeline (Top figure) can be transformed, by applying \texttt{@mpx.filter\_grad} to the forward pass. It automatically casts all inputs to half precision and calculates the gradient via dynamic loss scaling and \texttt{eqx.filter\_grad}.}}
%     \label{fig:overview}
% \end{figure}

@@

\subsection{Model Implementation}
\label{sec:eqxvsmpxmodelupdate}

% \begin{example}[t]
% \begin{pbox}[]{}
% \begin{minted}[linenos, breaklines, mathescape, fontsize=\footnotesize, xleftmargin=2em]{python}
% class MultiHeadAttentionBlock(eqx.Module):
%     dense_qs: eqx.nn.Linear
%     dense_ks: eqx.nn.Linear
%     dense_vs: eqx.nn.Linear
%     dense_o: eqx.nn.Linear
%     num_heads: int
%     layer_norm: eqx.nn.LayerNorm

%     def __init__(self, feature_dim, num_heads, key):
%         self.num_heads = num_heads
%         key, subkey = jax.random.split(key)
%         self.dense_qs = eqx.nn.Linear(
%             feature_dim, feature_dim, key=subkey)
%         # same for dense_ks, dense_vs, dense_o

%         self.layer_norm = eqx.nn.LayerNorm(feature_dim)

%     def attention(q, k, v):
%         attention_scores = q @ k.T / jnp.sqrt(q.shape[-1])
%         attention_scores = mpx.force_full_precision(
%             jax.nn.softmax, attention_scores.dtype)(attention_scores, axis=-1)
%         return attention_scores @ v

%     def __call__(self, inputs):
%         inputs_after_layernorm = jax.vmap(mpx.force_full_precision(
%             self.layer_norm, inputs.dtype))(inputs)
%         qs = jax.vmap(self.dense_qs)(inputs_after_layernorm)
%         qs = es.jax_einshape("n(hf)->hnf", qs, h=self.num_heads)
%         # same for ks and vs...

%         outputs = jax.vmap(self.attention, in_axes=(0, 0, 0))(qs, ks, vs)
%         outputs = es.jax_einshape("hnf->n(hf)", outputs)
%         outputs = jax.vmap(self.dense_o)(outputs)
%         outputs += inputs

%         return outputs
% \end{minted}
% \end{pbox}
% \caption{Implementation of multi-headed self-attention via MPX. \capt{For the largest part, the model can be implemented in common toolboxes (here, Equinox~\citep{kidger2021equinox}). Only critical function that should be implemented in full precision, like softmax or layernorm (due to mean and std calculations), have to be transformed via \mpx{}.}}
% \label{ex:model}
% \end{example}

For the largest part, the implementation of the model must not be changed.
As \mpx{} works with arbitrary PyTrees, every Toolbox that defines their model/parameters as PyTrees, like Flax~\citep{flax2020github} or Equinox~\citep{kidger2021equinox} can be used in conjunction with \mpx{} (Example~\ref{ex:model}). 
Only parts of the model that contain critical operations with a high risk of overflows, must be transformed to full precision via \texttt{mpx.force\_full\_precision}.
However, many jax and also equinox already implement many critical functions in a numerically stable way by casting to full precision if necessary, so that only few changes are necessary.

% \subsection{Model Updates}
% \label{sec:eqxvsmpxmodelupdate}
Model updates via \mpx{} closely resemble those performed with Equinox (Example~\ref{ex:update}). 
Instead of calling \texttt{eqx.filter\_grad}, \texttt{mpx.filter\_grad} is used. Similarly, \texttt{mpx.optimizer\_update} replaces invoking the Optax optimizer followed by a model update.
The remainder of the training step—such as loading batches, sharding the model and batch—remains unchanged. 
All these steps can be compiled using tools like \texttt{eqx.filter\_jit}.
